When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-15 doc/paper/intro.tex
intro rough done, onto body

The objective of the present manuscript is to use reasonable
approximations as a tool to develop intuition for plant response to
atmospheric drying and evaluate the VPD dependence of ET. This
intuition will aid interpretation of observations, full complexity
models, and facilitate the disentanglement of complex land-atmosphere
feedbacks. In the past, similar approaches were used to understand
interactions between stomatal conductance, evapotranspiration and the
environment \citep[e.g.,][]{Jarvis_1986, Mcnaughton_1991}. However, at
the time researchers' understanding of the form of VPD's effect on
plant physiology was limited, so they could not explore the
sensitivity of ET to VPD, including VPD's effect on stomatal
conductance and plant function.

Recent results have drastically improved our understanding of VPD's
impact on physiology, especially at the leaf
level. \citet{MEDLYN_2011} developed a model for leaf-scale stomatal
conductance ($g_s$), including VPD response, by combining an optimal
photosynthesis theory \citep{Cowan_1977} with an empirical approach,
and extended use of this model to the ecosystem scale in
\citet{Medlyn_2017}. Additionally, \citet{Zhou_2014} demonstrated that
a quantity underlying water use efficiency
$\left(uWUE = \frac{GPP\; \sqrt{VPD}}{ET}\right)$ properly captures a
constant relationship between GPP, ET, and VPD over a diurnal cycle at
the ecosystem scale. uWUE is also remarkably well conserved in the
growing season across space and time, within a PFT
\citep{Zhou_2015}. While stomatal conductance parameterizations and
uWUE greatly simplify complex plant physiological processes, they
still capture leading order ecosystem behavior for vegetated
surfaces \citep{Medlyn_2017, Zhou_2014}, and are novel tools to
transparently develop intuition for the behavior of complex, multiscale
ecohydrologic systems.

In this manuscript,
we leverage uWUE and recent developments in stomatal conductance
parameterizations \citep{MEDLYN_2011} to derive the theoretical
one-way response of ET to VPD with other environmental variables
properly controlled for, i.e. we develop a framework for evaluating
the partial derivative of ET with respect to VPD. For the first time,
we explicitly include VPD's full effect on stomatal conductance,
including its leading order impact on photosynthesis. Our theory is
validated and tested at multiple eddy-covariance stations spanning
various climates and plant functional types.

>>>>

The objective of the present manuscript is to use reasonable
approximations established in prior research as a tool to develop a
framework for understanding plant response to atmospheric drying and
evaluating the VPD dependence of ET. This framework will aid
interpretation of observations, full complexity models, and facilitate
the disentanglement of complex land-atmosphere feedbacks. In the past,
similar approaches were used to understand interactions between
stomatal conductance, evapotranspiration and the environment
\citep[e.g.,][]{Jarvis_1986, Mcnaughton_1991}. However, at the time
researchers' understanding of the form of VPD's effect on plant
physiology was limited, so they could not explore the sensitivity of
ET to VPD, including VPD's effect on stomatal conductance and plant
function.

Recent results have drastically improved our understanding of VPD's
impact on physiology, especially at the leaf
level. \citet{MEDLYN_2011} developed a model for leaf-scale stomatal
conductance ($g_s$), including VPD response, by combining an optimal
photosynthesis theory \citep{Cowan_1977} with an empirical approach,
and extended use of this model to the ecosystem scale in
\citet{Medlyn_2017}. Additionally, \citet{Zhou_2014} demonstrated that
a quantity underlying water use efficiency $\left(uWUE = \frac{GPP\;
\sqrt{VPD}}{ET}\right)$ properly captures a constant relationship
between GPP, ET, and VPD over a diurnal cycle at the ecosystem
scale. uWUE is also relatively well conserved in the growing season
across space and time, within a PFT \citep{Zhou_2015}. While stomatal
conductance parameterizations and uWUE greatly simplify complex plant
physiological processes, they still capture ecosystem behavior for
vegetated surfaces \citep{Medlyn_2017, Zhou_2014}, and are novel tools
to transparently develop intuition for the behavior of complex,
multiscale ecohydrologic systems.

In this manuscript, we leverage uWUE and recent developments in
stomatal conductance parameterizations \citep{MEDLYN_2011} to derive
the theoretical one-way response of ET to VPD with other environmental
variables properly controlled for, i.e. we develop a framework for
evaluating the partial derivative of ET with respect to VPD. For the
first time, we explicitly include VPD's full effect on stomatal
conductance, including its impact on photosynthesis. We explore the
range of possible ET responses to VPD, given parameters previously
established in peer reviewed literature. Additionally, we show the
sensitivity of the ET-VPD relationship to model and framework choice,
highlighting the importance of: 1) future research on stomatal
conductance and ecosystem scaling, and 2) thoughtful selection of
photosynthesis and stomatal conductance model in more sophisticated
land surface and earth system models.
